---
title: "DS 1300 - Final Project"
author: "Brock Wyma, Conner Woods, Emmanuel Garcia"
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-line-numbers: true
    code-summary: "SHOW ME THE CODE!"
    code-tools: true
    theme: yeti
editor: visual
---

# Introduction

The City of Longbeach Animal Care Services rescues and shelters animals in need of assistance. Their operation aims to support the people and animals of their community through outreach, education, guidance, and support services. Several core values guide them through business decisions. These values include compassion for others, passion to achieve their vision, dedication to their craft, and ethical decision making. While these core values create a foundation for Longbeach Animal Care Services, partnerships, animal well-being, and data specifically guide company decisions. Given two data sets from Longbeach Animal Care Services, this report aims to analyse the data and extract readable information to help the Longbeach Animal Services achieve their goals.

```{r, message=FALSE, warning=FALSE, results='hide'}
# Load Libraries
library(tidyverse)
library(openxlsx)
library(DT)
library(gt)
library(lubridate)
library(leaflet)
library(sf)
library(viridis)

# Read in the data set
longbeach <- read.csv("Data/longbeach.csv")

# Type convert works well on this data set
longbeach <- longbeach %>%
  as_tibble() %>%
  type_convert()

# Read in data set descriptions
longbeach_dd <- read.xlsx("Data/longbeach_dd.xlsx", sheet = 1)

# Showing the data for short introduction to the data set
summary(longbeach)
print(longbeach_dd)
```

The data contains 29,787 entries/rows of 22 variables/columns. Each row contains one case on an animal which was involved with the shelter at some point in time. The columns describe components of the case such as the animals species, the way it came into the shelter, the way it left the shelter, and many more descriptions.

From this initial inspection of the data being given to us, we raise the following research questions which will be answered later in this report:

1.  Are there any seasonal patterns related to intake type and count?

2.  Are there any outcomes that occur more commonly than others and does the animal species affect the outcome?

3.  How has the number of pet adoptions changed over time?

4.  What types of animals are adopted? Do certain animal characteristics influence the chances of adoption for an animal?

5.  Are there locations of greater intake? If so, is it due to the animals species?

# Methods

## A Quick Deep Dive

```{r, message=FALSE, warning=FALSE, results='hide'}
# Providing a summary of the dates
summary(longbeach$outcome_date)
summary(longbeach$intake_date)

# Finding the number of unique column values for outcome type
n_distinct(longbeach$outcome_type)
unique(longbeach$outcome_type)

# NA counts for each variable
longbeach_NA_count <- longbeach %>%
  summarise(across(everything(), ~sum(is.na(.))))
```

The target variable is outcome type with outcome sub-type and outcome date providing additional details of the outcome. With 19 unique cases, outcomes range from the worst being death or euthanasia to the animal being returned to its owner or adopted. Some variables that might be of interest include the animals name and species. Animals with names might be more likely to end up back with their owners, and certain types of animals may be more likely to be adopted. The intake condition, type, and sub-type should also be useful for analyzing how and why animals arrive at the shelter. They also provide several variables describing the location of an animals intake such as crossing, latitude, longitude, and geopoint, allowing us to look into the affect of location on the other variables.

Looking into the values for some of the data, animal name contains 12,161 missing values and secondary color contains 15,604 missing values. This may make it harder to analyze extract useful information from these variables. The same can be said for date of birth and outcome sub-type with \~3,500 missing entries. The reason for intake variable contains 27,784 missing values (\~93% of the data) making it almost useless for our analysis. The other variables either contain no missing values or a very small amount of missing data. Analyzing the intake and outcome date variables, the data ranges from January 1, 2017 to December 31, 2024 which is about 8 years of data. This gives a us a very generous range of time which allows us to effectively analyze the data over time.

## Cleaning the Data (Fixing Missing Data and Extreme Values)

```{r, message=FALSE, warning=FALSE, results='hide'}
# Removing variables with lots of missing dtaa
longbeach <- longbeach %>%
  select(-reason_for_intake, -jurisdiction, -sex, -dob, -geopoint)
```

To start cleaning the data, we removed unnecessary rows or columns. Since the reason for intake contains too much missing data and doesn't help answer our questions, the column was removed. Similarly, the jurisdiction, sex, and date of birth variables don't help answer our questions, so they were removed. The geopoint column simply combines the latitude and longitude columns into one value. In order to remove redundant data, we kept the more informative columns (latitude, longitude) and remove geopoint.

```{r, message=FALSE, warning=FALSE, results='hide', fig.show='hide'}
# Boxplots to look for outliers in the intake/outcome dates
ggplot(longbeach) +
  geom_boxplot(aes(x = "Intake Date", y = intake_date), fill = "skyblue") +
  geom_boxplot(aes(x = "Outcome Date", y = outcome_date), fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Boxplots for Intake/Outcome Dates", x = "Variable", y = "Date")

# 5 Number Summaries for latitude and longitude
summary(longbeach$latitude)
summary(longbeach$longitude)

# Removing the outliers using the 1.5*IQR Rule
longbeach <- longbeach %>%
  filter(
    between(latitude, quantile(latitude, 0.25) - 1.5 * IQR(latitude), quantile(latitude, 0.75) + 1.5 * IQR(latitude)),
    between(longitude, quantile(longitude, 0.25) - 1.5 * IQR(longitude), quantile(longitude, 0.75) + 1.5 * IQR(longitude))
  )
```

Since most of the data is categorical, there are only a few places that extreme values could exist. Outliers could appear in the date variables or the latitude/longitude variables. For the intake and outcome dates, a boxplot shows no outliers. For the latitude and longitude variables, looking at the 5-number summary, there appeared to be outliers due to the massive gape between the min/max and the quartiles. Using the 1.5\*IQR Rule, we removed these outliers. This removed 214 rows of data or 0.72% of the data, taking our total entries down to 29,573.

## Classifying Data Types

Here, we classified the variables of interest in the data into their respective types:

| Variable             | Type                    |
|----------------------|-------------------------|
| Intake Date          | Categorical Ordinal     |
| Type of Intake       | Categorical Nominal     |
| Type of Outcome      | Categorical Nominal     |
| Animal Species       | Categorical Nominal     |
| Animal Name          | Categorical Nominal     |
| Latitude of Intake   | Nominal Continuous      |
| Longitude of Intake  | Nominal Continuous      |

This showed what kind of visual and numeric summaries can be used on the data. Since we mostly have categorical variables, we shifted most of our analysis to various categorical summaries on one or more variables. These include numeric summaries such as frequency tables and contingency tables as well as visual summaries such as bar plots. There was also access to visual summaries such as maps from the latitude/longitude variables and time series plots from the date variable.

## Summarizing the Data

For summarizing the seasonal changes for animal intake, we grouped the data by months. For the first visual summary, we grouped the data by intake type and counted the average intake over the 8 years for each month. We took the top three most common intake types and plotted those on a time series plot. For the second visual summary, we also made a time series plot, however, we disregard the intake type in an attempt to get a better visualization of the effect of time on intake. We split the data into two year chunks and averaged the total intakes per month, plotting each chunk as a separate line. 

Taking the outcome types, we created a bar chart for this categorical variable. We sorted the bar chart in ascending order. We then created a table containing the top three outcomes for each animal type. 

For adoption rates, we added up all of the adoption related outcomes for each year of data. We plotted these outcomes on a yearly time series plot for all eight years of data.

Taking a deeper look into adoptions, we created a contingency table of animals with a name/without a name and adopted/not adopted animals in order to analyze any effect of animal names on adoptions. We then created a bar chart showing the percentage of adoptions for each animal type. Since this chart only paints a part of the story, we created a second bar chart along side it. This bar chart contains the percentages that each animal type represents among only adopted animals. 

We created a leaflet map to visualize the intake counts relative to location. The map was marked with gridlike clusters of varying color to signify areas of higher or lower intake. It contained an interactive filter on animal type to allow the user to pick which animal species to show on the map. 

# Results

## Seasonal Patters in Relation to Animal Intake

```{r}
# Find the top three most common intake types
top_types <- longbeach %>%
  count(intake_type, sort = TRUE) %>%
  slice_head(n = 3) %>%
  pull(intake_type)

# Prepare the data for plotting...
plot_data <- longbeach %>%
  filter(intake_type %in% top_types, !is.na(intake_date)) %>%
  mutate(month_num = as.integer(format(intake_date, "%m"))) %>%
  group_by(intake_type, year = format(intake_date, "%Y"), month_num) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(intake_type, month_num) %>%
  summarise(avg_total = mean(count), .groups = "drop")

# Plot the time series plot
ggplot(plot_data, aes(x = month_num, y = avg_total, color = intake_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Figure 1: Average Monthly Intakes for Top 3 Intake Types",
    x = "Month",
    y = "Average Intake Count",
    color = "Intake Type"
  ) +
  theme_bw()
```

The results of Figure 1 showed that the most common intake type over the entire year on average was stray animals. Stray animals peaked at June with an average count of 300. Stray animals counts also fell to low counts around November to March. Wildlife, the second most common intake type, shows a similar trend to the stray counts with a lower numbers during November to February. Owners surrendering their animal had the lowest of the top three intake types with relatively even counts during the year.

```{r}
# Create a summary subset of the data
seasonal_summary <- longbeach %>%
  mutate(month = as.Date(paste0(format(intake_date, "%Y-%m"), "-01"))) %>%
  group_by(month) %>%
  summarise(
    intake_count = n(),
    .groups = "drop"
  )

# A time series plot to visualize the intake number per month better.
plot_data <- seasonal_summary %>%
  mutate(
    year = as.integer(format(month, "%Y")),
    month_num = as.integer(format(month, "%m")),
    period = case_when(
      year %in% 2017:2018 ~ "2017–2018",
      year %in% 2019:2020 ~ "2019-2020",
      year %in% 2021:2022 ~ "2021-2022",
      year %in% 2023:2024 ~ "2023–2024",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(period, month_num) %>%
  summarise(
    avg_total = mean(intake_count),
    .groups = "drop"
  )

ggplot(plot_data, aes(x = month_num, y = avg_total, color = period, group = period)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Figure 2: Average Monthly Intakes (2 Year Intervals)",
    x = "Month",
    y = "Average Intake Count",
    color = "Period of Time"
  ) +
  theme_bw()
```

Intake counts were higher during the months of April to October in Figure 2. The peak intake counts for 2017-2018, 2019-2020, 2021-2022, and 2023-2024 respectively were May/August with 491.5, June with 464, September with 339, and May with 524.

## Animal Outcomes

```{r}
# Plotting the bar chart for outcome types
longbeach %>%
  filter(!is.na(outcome_type)) %>%
  ggplot(aes(y = fct_rev(fct_infreq(outcome_type)))) +
  geom_bar(fill = "steelblue") +
  labs(y = "Outcome Type",
       x = "Count",
       title = "Figure 3: Number of Occurences for the Different Outcomes") +
  theme_bw()
```

From Figure 3, Rescue, adoption, euthanasia, transfer, and returned to owner had a much higher count than all of the other outcomes with rescue being the highest.

```{r}
# Grabbing the top 3 groupings per animal type for outcome types
varyBySpecies <- longbeach |>
  group_by(animal_type, outcome_type) |>
  summarize(count = n(), .groups = "drop") |>
  group_by(animal_type) |>
  slice_max(order_by = count, n = 3)

# Displaying the information as a table
datatable(varyBySpecies, 
          caption = "Table 1: Count of Outcomes per Species",
          colnames = c("Species", "Outcome Type", "Count"),
          filter = "top")
```

Table 1 shows a subset of the top three outcome types per animal species. For animals with less than three outcome types, all of their outcome types are shown in the table. Using Table2, Cats had the highest rescue rate, and dogs had the highest rate of being returned to their owner. A large number of cats and dogs get adopted with it being the second highest outcome for each animal. Wild animals had a very high euthanasia rate compared to its other outcomes.

## Pet Adoption Rates

```{r}
#subset of the data that only includes adoptions
adopted <- longbeach |>
  filter(outcome_type == "adoption" | outcome_type == "foster to adopt") |>
  mutate(year = format(outcome_date, "%Y")) |>
  group_by(year) |>
  summarize(count = n()) 
  
# Time series plot of adoption counts per year over the 8 years of data. 
ggplot(adopted, aes(x = year, y = count, group = 1)) +
  geom_line() +
  geom_point() +
  labs(x = "Intake Date",
       y = "Number of Adoptions",
       title = "Figure 4: Number of Adoptions Over Time Based on Intake Date") +
  theme_bw()
```

From Figure 4, there was a local spike in adoptions during 2019. Ever since 2021, adoptions increased over the years with 1463 adoptions during 2024. The biggest increase was from 2021 to 2022 with an increase of 723 adoptions.

## Types of Pets Adopted

```{r}
adoption_table <- longbeach %>%
  mutate(
    Name = ifelse(is.na(animal_name), "Doesn't Have a Name", "Has a Name"),
    Adopted = ifelse(outcome_type %in% c("adoption", "foster to adopt"), "Adopted", "Not Adopted")
  ) %>%
  count(Name, Adopted) %>%
  tidyr::pivot_wider(names_from = Adopted, values_from = n, values_fill = 0)

# Display as a styled table
gt(adoption_table) %>%
  tab_header(
    title = "Table 2: Adoption Outcome by Animal Names",
    subtitle = "Comparing adoption rates between animals with and without names"
  ) %>%
  cols_label(
    Name = "Animal Name Status",
    Adopted = "Adopted",
    `Not Adopted` = "Not Adopted"
  ) %>%
  fmt_number(columns = 2:3, decimals = 0, sep_mark = ",") %>%
  tab_options(
    table.border.top.color = "gray",
    table.border.bottom.color = "gray"
  )
```
Using Table 2, out of animals which were adopted, 97% of them had names. On the other hand, 43.4% of the animals which were not adopted had names.

```{r}
# Calculate the adoption rates per animal type
adoption_rates <- longbeach %>%
  mutate(adopted = outcome_type %in% c("adoption", "foster to adopt")) %>%
  group_by(animal_type) %>%
  summarise(
    total = n(),
    adopted_count = sum(adopted),
    adoption_rate = round(100 * adopted_count / total, 2)
  ) 

ggplot(adoption_rates, aes(x = adoption_rate, y = reorder(animal_type, adoption_rate))) +
  geom_col(fill = "#69b3a2") +
  scale_x_continuous(limits = c(0, 100), 
                     breaks = seq(0, 100, by = 10),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title = "Figure 5: Adoption Rate by Animal Type",
    x = "Adoption Rate (%)",
    y = "Animal Type"
  ) +
  theme_bw()
```
Figure 5 shows that livestock had the highest adoption rate at 60%. The next highest rate were guinea pigs with a 33.72% adoption rate. Rabbits, dogs, and cats were all adopted at similar rates around ~27% of the time. Wild animals, with a total of 1412 entries, were never adopted. Amphibians, with a total of 3 entreis, were never adopted.

```{r}
# Filter to adopted animals and calculate % by animal type
adoption_distribution <- longbeach %>%
  filter(outcome_type %in% c("adoption", "foster to adopt")) %>%
  mutate(animal_type = str_to_title(animal_type)) %>%
  count(animal_type) %>%
  mutate(
    percent = n / sum(n) * 100
  )

# Plot as a horizontal bar chart
ggplot(adoption_distribution, aes(x = percent, y = reorder(animal_type, percent))) +
  geom_col(fill = "#69b3a2") +
  labs(
    title = "Figure 6: Adoption Distribution by Animal Type",
    x = "Percentage of Adoptions",
    y = "Animal Type"
  ) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  theme_bw()
```
From Figure 6, cats and dogs had the highest amount of adoptions compared to all adoptions from all animals. Cats and dogs combined also account for almost all adoptions at the Longbeach Animal Care Services.

## Animal Intake According to Location

```{r}
# Convert to sf and filter valid coordinates
df_sf <- longbeach %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  mutate(animal_type = tools::toTitleCase(animal_type)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Transform to meters for square grid
df_sf_m <- st_transform(df_sf, crs = 3857)

# Create square grid (1000m x 1000m)
cell_size <- 1000
grid <- st_make_grid(df_sf_m, cellsize = cell_size, square = TRUE)

# Convert grid to sf and join with points
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)
joined <- st_join(df_sf_m, grid_sf)

# Count animals by grid and type
grid_counts <- joined %>%
  st_drop_geometry() %>%
  count(grid_id, animal_type, name = "count") %>%
  left_join(grid_sf, by = "grid_id") %>%
  st_as_sf()

# Count all animals regardless of type
grid_total <- joined %>%
  st_drop_geometry() %>%
  count(grid_id, name = "count") %>%
  left_join(grid_sf, by = "grid_id") %>%
  mutate(animal_type = "All Animals") %>%
  st_as_sf()

# Combine specific animal types with "All Animals" section
grid_combined <- rbind(grid_counts, grid_total)

# Transform back to WGS84 for leaflet
grid_combined_wgs <- st_transform(grid_combined, crs = 4326)

# Setup map with layers per animal type
pal <- colorNumeric("viridis", domain = grid_combined_wgs$count)
animal_types <- unique(grid_combined_wgs$animal_type)

m <- leaflet() %>%
  addProviderTiles("CartoDB.Positron")

for (type in animal_types) {
  m <- m %>%
    addPolygons(
      data = filter(grid_combined_wgs, animal_type == type),
      fillColor = ~pal(count),
      fillOpacity = 0.5,
      color = "#444444",
      weight = 1,
      popup = ~paste("Animal: ", animal_type, "<br>Count:", count),
      group = type
    )
}

m %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = grid_combined_wgs$count,
    title = "Intake Count"
  ) %>%
  addLayersControl(
    baseGroups = animal_types,
    options = layersControlOptions(collapsed = FALSE)
  )
```

The map shows that the Longbeach shelter intakes animals, regardless of their species, around similar locations locations. Looking at the intake for all animals, there tends to be higher concentrations of intake around Longbeach itself, North Longbeach, and the Los Alamitos area.

# Discussion 

One of the clearest trends observed is that most animals are taken into the shelter during summer months. In particular, from April to July, peaking in May. There was a decrease in the intake of animals to the shelter during the years of 2020 and 2021 most likely due to the workers, or lack of workers, ability to bring the animals into the shelter due to the COVID-19 pandemic. We found that the most common type of animal taken into the shelter are strays. This is because the other categories are less common. To reduce strays we can inform the public about the consequences of abandoning animals. Fixing and neutering strays on the street would help reduce the stray population. And lastly increasing shelter funding can help the shelter obtain more animals from the streets by hiring more employees and having the resources to house the animals. 

Unfortunately, euthanasia is the third most common outcome for the animals brought into the shelter. Cats, in particular, see the outcome far too much. Wild animals rarely get adopted and require distinct management strategies, so one option would be to prioritize the resources and space for animals that may have a chance of being adopted. This is a very concerning outcome, thus increasing community awareness is crucial to combat this issue. The shelter could utilize social media to better spread awareness about the excess intake of animals. In that, the shelters need people to adopt the pets urgently, and, for those who can't afford to adopt, volunteering could be useful. Reducing adoption fees can also decrease the euthanasia outcome which the shelter currently deals with. 

The adoption rates of named animals are significantly greater than those of unidentified animals, which is another intriguing behavioral finding. This could be because people believe that named animals are more socialized since they have lived in previous households. Thus, Longbeach can put more adoption effort towards named animals first since they have a higher success rate of being adopted. They could also give names to the animals up for adoption. Pet adoptions rates steadily rise from 2017-2019 from 350 to about 625, but then they drop off, most likely due to Covid, during 2020-2021. Around 2020 pet adoption rates reached an all time low of zero. However, after 2021, they continue to rise until today to a maximum of around 1375. Under normal conditions, we can probably expect this trend to continue until it reaches a steady state. The fact that wild animals are almost never adopted further supports the idea that domestic pets should be the main focus of adoption campaigns, whereas rehabilitation, relocation, or other non-adoption outcomes should be the focus for wild animals taken in by the shelter.

Knowing and being able to predict animal intake location allows Longbeach to allocate the proper amount of resources to the correct locations. While the map failed to show any real differences in intake location for different animal species, it shows a good trend of where animals get found and taken to the shelter. We can see significant peaks of animal intake in the central Lonbeach, North Longbeach, and Los Alamitos areas. Allocating more resources to these areas should ease up the intake process, making the animal intake process run much smoother. This information can also help Longbeach implement preventative measures. Since Longbeach aims to build a solid community of people and animals, they can put more work into these communities with higher animal abandonment.

Overall, these trends in the data can offer the shelter clear guidance for making future plans. It is crucial for the shelter to be prepared for the seasonal intake spikes during the summer months through hiring staffing, having proper resources, and creating foster campaigns in advance. To cut down on strays, expand owner education and spay/neuter programs. It is important for cats and other suitable domestic animals to be the focus of euthanasia reduction initiatives. To increase the likelihood that an animal will be adopted, encourage naming of animals upon their intake. In order for us to conserve resources and space in the shelter, it would be a good idea to shift the emphasis of wild animals and find them an appropriate alternative. By the shelter aligning with the strategies suggested previously, they will be able to make more effective decisions and potentially improve outcomes while optimizing their resource use.

During our data preparation, many columns and a few rows of data were removed or unused. The number of rows removed amounted to such a small percent of the data that it most likely had no effect on the results. On the other hand, looking into alternative research questions that make use of the unused columns would be a great next step. It could be useful to look into more details on the types of intakes and outcomes or the effect of an animals color on certain outcomes. This report took a deep dive into the adoption outcome which leaves the other outcomes for future research. While we covered some of the more important aspects of the data in this report, we hope to look into the multiple unexplored avenues in the data in the future.
